Reducing RES Droughts through the integration of wind and PV

### 2024-12-09 Article/main.tex
Update Article

The EU aims to generate at least 69\% of its electricity from renewable energy sources (RES) by 2030, up from 41\% in 2022 \cite{eurostat2023share}. While this transition is essential for reducing greenhouse gas emissions, it also highlights the challenge of managing the variability of weather-dependent energy sources like wind and photovoltaic (PV) power. This challenge is compounded by the increasing electrification of energy sectors, which places greater demand on the power system and makes it more sensitive to meteorological conditions~\cite{bloomfield2021, bloomfield2016, vanderWiel2019drought}. Periods of low renewable generation, known as \textit{Dunkelflaute} or RES droughts, pose significant risks to grid stability and energy security, emphasizing the need for a resilient energy system to meet both growing electricity demand and decarbonization targets.

@@

The C3S-E dataset estimates wind energy using wind speeds at 100 metres ($u_{100}$, $v_{100}$) and a standard turbine model, the Vestas V136/3450, with a fixed hub height of 100 m. This choice is based on expert advice and the trend in wind turbine installation. The PV generation model used by C3S-E uses two ERA5 variables: Surface solar radiation downwards ($ssrd$) and air temperature ($t2m$). PV generation is calculated multiple times, using the same model with different azimuth and tilt angles. The results are aggregated based on a statistical distribution of the module angles based on the geographical location~\cite{saintdrenan2018solar}.

@@

In addition to analysing wind and PV generation separately, a combined CF was computed for each model by averaging wind and PV generation, weighted by their installed capacities at the end of 2023 (6.7 GW for wind and 0.6 GW for PV). Given that PV capacity in Ireland is low in 2023, and to explore how a more balanced distribution of wind and PV might impact RES droughts, this study also considered a balanced scenario, where the installed PV capacity is assumed to increase to 8 GW, while wind capacity rises to 9 GW. These values are based on targets outlined in the roadmap published by EirGrid and SONI~\cite{eirgrid2023future}. New time series were generated for both the Atlite and C3S-E G PV models, incorporating a revised distribution of installed capacity across Ireland as specified in the roadmap. For wind, the CF time series remains unchanged, as significant shifts in the location of wind farms are not expected. The PV and wind time series were then combined using updated weights that reflect the installed capacities in the balanced scenario (9 GW for wind and 8 GW for PV).

@@

The Atlite model allows the user to select certain PV panel characteristics. In this study, the three PV panel types available in the Atlite model were considered. Following the same methodology as in the previous section, the three available models were compared using four skill scores (CC, RMSE, MB, and area under the curve). Based on the best-performing metrics, the Breyer PV panel model was selected \cite{beyer2004pv}, using the Kaneka Hybrid panel option. For all PV farm locations, the azimuth angle is fixed at 180\textdegree (due south), and the optimal tilt angle option is applied.

@@

\begin{figure}[!ht]
	\centering
	\includegraphics{verification_pv_number_events.pdf}
	\caption{Annual number of PV drought events for Atlite (red), C3S-E G (blue), and C3S-E N (purple) and EirGrid (black outline) in 2023}
	\label{fig:bar_number_events_verification_pv}
\end{figure}

@@

The first part of the analysis examines the annual number of RES drought events across the three datasets. For wind (Fig.~\ref{fig:boxplot_number_events}a), the number of events decreases as the duration range increases, with very few events lasting more than seven days. For PV (Fig.~\ref{fig:boxplot_number_events}b), the number of events also declines as the duration range extends from one to eight days, followed by a slight increase for longer durations. This increase is due to extended low-generation periods occurring from November to March, depending on the dataset. When comparing wind and PV results (Fig.~\ref{fig:boxplot_number_events}a \& Fig.~\ref{fig:boxplot_number_events}b), the median, first, and third quartiles for PV are consistently higher or equal to those for wind, across all duration ranges and datasets. This is due to the typically lower CF of PV compared to wind, especially in a region like Ireland where solar potential is limited. PV generation is also zero at night and constrained by the daily solar cycle, leading to a naturally higher frequency of RES droughts in PV compared to wind.

@@

For wind (Fig.~\ref{fig:res_droughts_seasonality}a), RES drought percentages are higher in summer than in winter. In the Atlite dataset, for instance, an average of 24\% of hours in summer (JJA) are identified as RES droughts, compared to only 4\% in winter (DJF). This seasonal variation is influenced by the wind power curve model used to estimate CF, where the shape of the curve in lower wind speed regions (3-10 m/s) leads to significant differences in CF under low wind conditions. In contrast, the results for PV (Fig.~\ref{fig:res_droughts_seasonality}b) show a higher percentage in winter, with RES droughts occurring over 60\% of the time regardless of the dataset. The Atlite results show a higher percentage of RES drought hours for wind, and a slightly lower percentage for PV, compared to the two C3S-E datasets.

@@

\newpage
\section*{Appendix}

>>>>

The EU aims to generate at least 69\% of its electricity from renewable energy sources (RES) by 2030, up from 41\% in 2022 \cite{eurostat2023share}. While this transition is essential for reducing greenhouse gas emissions, it also highlights the challenge of managing the variability of weather-dependent energy sources such as wind and photovoltaic (PV) power. This challenge is compounded by the increasing electrification of energy sectors, which places greater demand on the power system and makes it more sensitive to meteorological conditions~\cite{bloomfield2021, bloomfield2016, vanderWiel2019drought}. Periods of low renewable generation, known as \textit{Dunkelflaute} or RES droughts, pose significant risks to system adequacy and energy security, emphasizing the need for a resilient energy system to meet both growing electricity demand and decarbonization targets.

@@

The C3S-E dataset estimates wind energy using wind speeds at 100 metres ($u_{100}$, $v_{100}$) and a standard turbine model, the Vestas V136/3450, with a fixed hub height of 100 meters. This choice is based on expert advice and the trend in wind turbine installation. The PV generation model used by C3S-E uses two ERA5 variables: Surface solar radiation downwards ($ssrd$) and air temperature ($t2m$). PV generation is calculated multiple times, using the same model with different azimuth and tilt angles. The results are aggregated based on a statistical distribution of the module angles based on the geographical location~\cite{saintdrenan2018solar}.

@@

In addition to analysing wind and PV generation separately, a combined CF was computed for each model by averaging wind and PV generation, weighted by their installed capacities at the end of 2023 (6.7 GW for wind and 0.6 GW for PV). Given that PV capacity in Ireland is low in 2023, and to explore how a more balanced distribution of wind and PV might impact RES droughts, this study also considered a balanced scenario, where the installed PV capacity is assumed to increase to 8 GW, while wind capacity rises to 9 GW. These values are based on targets outlined in the roadmap published by the 2024 Climate Action Plan~\cite{cap2024future}. New time series were generated for both the Atlite and C3S-E G PV models, incorporating a revised distribution of installed capacity across Ireland as specified in the roadmap. For wind, the CF time series remains unchanged, as significant shifts in the location of wind farms are not expected. The PV and wind time series were then combined using updated weights that reflect the installed capacities in the balanced scenario (9 GW for wind and 8 GW for PV).

@@

The Atlite model allows the user to select certain PV panel characteristics. In this study, the three PV panel types available in the Atlite model were considered (CSi, CdTe, Kanena). Following the same methodology as in the previous section, the three available models were compared using four skill scores (CC, RMSE, MB, and area under the curve). Based on the best-performing metrics, the Breyer PV panel model was selected \cite{beyer2004pv}, using the Kanena Hybrid panel option. For all PV farm locations, the azimuth angle is fixed at 180\textdegree (due south), and the optimal tilt angle option is applied.

@@

\begin{figure}[!ht]
	\centering
	\includegraphics{verification_pv_number_events.pdf}
	\caption{Number of PV drought events for Atlite (red), C3S-E G (blue), and C3S-E N (purple) and EirGrid (black outline) during 2023}
	\label{fig:bar_number_events_verification_pv}
\end{figure}

@@

The first part of the analysis examines the annual number of RES drought events across the three datasets. For wind (Fig.~\ref{fig:boxplot_number_events}a), the number of events decreases as the duration range increases, with very few events lasting more than seven days. For PV (Fig.~\ref{fig:boxplot_number_events}b), the number of events also declines as the duration range extends from one to eight days, followed by a slight increase for longer durations. This increase is due to extended low-generation periods occurring from November to March, depending on the dataset. When comparing wind and PV results (Fig.~\ref{fig:boxplot_number_events}a \& Fig.~\ref{fig:boxplot_number_events}b), the median, first, and third quartiles for PV are consistently higher or equal to those for wind, across all duration ranges and datasets. This is due to the typically lower CF of PV compared to wind, especially in a region such as Ireland where solar potential is limited. PV generation is also zero at night and constrained by the daily solar cycle, leading to a naturally higher frequency of RES droughts in PV compared to wind.

@@

For wind (Fig.~\ref{fig:res_droughts_seasonality}a), RES drought percentages are higher in summer than in winter. In the Atlite dataset, for instance, an average of 24\% of hours in summer (June-July-August) are identified as RES droughts, compared to only 4\% in winter (December-January-February). This seasonal variation is influenced by the wind power curve model used to estimate CF, where the shape of the curve in lower wind speed regions (3-10 m/s) leads to significant differences in CF under low wind conditions. In contrast, the results for PV (Fig.~\ref{fig:res_droughts_seasonality}b) show a higher percentage in winter, with RES droughts occurring over 60\% of the time regardless of the dataset. The Atlite results show a higher percentage of RES drought hours for wind, and a slightly lower percentage for PV, compared to the two C3S-E datasets.

@@
